---
title: "Monitoring One Series with the Multimodel Q Chart"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Monitoring One Series with the Multimodel Q Chart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette explains the QMM method implemented by `fromage`. It assumes
basic knowledge of R, autocorrelation, and ARMA models.

## The monitoring problem

We observe one temporal series. Its first $T$ observations form the
**reference sample (Phase I)**: observations collected while the process is
believed to be in control. Later observations form the **monitoring period
(Phase II)** and are checked in their arrival order.

The method is intended to detect changes in serial dependence. A signal means
that the residual autocorrelation in the current monitoring window is unusually
large compared with bootstrap series generated from the reference sample. A
signal does not identify the physical cause of the change.

Autocorrelation is relevant whenever observations are collected in time order
and nearby measurements are related. An autocorrelation plot is a standard way
to inspect that relationship [@nistAutocorrelation]. QMM goes further by
monitoring whether the fitted dependence structure remains adequate.

## A simulated process-monitoring case

Suppose a sensor records one measurement from a continuous process every
minute. Engineers have centered the measurements at the process target and
scaled them, so zero means "on target." Consecutive values remain correlated
because the process and its controller have memory.

We simulate 100 reference observations. The controller is then changed, after
which 40 new observations are collected. The change increases the ARMA
coefficients from `(0.4, 0.3)` to `(0.8, 0.7)`. It does not deliberately add a
mean shift. This is a realistic *scenario*, not a real dataset: knowing the
data-generating mechanism lets us demonstrate whether the chart responds to a
change in serial dependence.

```{r simulate}
library(fromage)

simulated <- simulate_arma_change(
  reference_length = 100,
  monitoring_length = 40,
  reference_coefficients = c(ar1 = 0.4, ma1 = 0.3),
  monitoring_coefficients = c(ar1 = 0.8, ma1 = 0.7),
  order = c(1, 1),
  burn_in = 1000,
  seed = 2026
)

plot(
  simulated$series,
  type = "l",
  xlab = "Observation number",
  ylab = "Standardized deviation from target",
  main = "Simulated process sensor"
)
abline(v = 100.5, lty = 2, col = "grey40")
legend(
  "topleft",
  legend = c("Observed series", "Start of Phase II"),
  lty = c(1, 2),
  col = c("black", "grey40"),
  bty = "n"
)
```

The package receives only the observed series and the Phase I boundary. It
does not receive the true simulation coefficients or innovations.

## Step 1: specify candidate ARMA models

It is often uncertain which ARMA order adequately describes a short reference
sample. The method therefore fits a set of candidate models instead of selecting
one model and ignoring the others.

```{r candidates}
candidates <- arma_candidates(max_p = 1, max_q = 1)
candidates
```

This convenience grid contains ARMA(1,0), ARMA(0,1), and ARMA(1,1). In a real
study, the candidate set should be chosen using knowledge about the process and
the purpose of the analysis. Here the short-memory set is reasonable because
each model uses at most one autoregressive term and one moving-average term,
and it includes the ARMA(1,1) mechanism used to generate the example. It is not
automatically better to fit every model the computer can handle
[@burnham2002].

## Step 2: fit and weight the models

For candidate model $m$, let $IC_m$ be its information-criterion value. The
difference from the best value in the candidate set is

$$
\Delta_m = IC_m - \min_j IC_j.
$$

The normalized weight is

$$
w_m = \frac{\exp(-\Delta_m/2)}
{\sum_j \exp(-\Delta_j/2)}.
$$

These weights describe relative support within this candidate set and for this
reference sample. They are not permanent probabilities that a model is true.
AICc is the default because it adds a finite-sample correction to AIC
[@hurvich1989; @burnham2004].

```{r fit}
fit <- fit_qmm(simulated$reference, candidates)
fit
```

Every candidate remains visible in `summary(fit)`, including a model that
failed or was excluded. The weights of the included models sum to one.

```{r fit-details}
summary(fit)[, c("model", "AICc", "delta", "weight", "included")]
```

In this realization, ARMA(1,0) and ARMA(1,1) both receive substantial support.
The method preserves that uncertainty instead of acting as though one selected
order were certainly correct.

## Step 3: calculate the multimodel Q statistic

For each retained model, the package filters the current window while keeping
the reference-sample coefficients fixed. It calculates the Ljung-Box statistic
$Q_m(L)$ from that model's residuals [@ljung1978]. The multimodel statistic at
horizon $L$ is

$$
Q_{MM}(L) = \sum_m w_m Q_m(L).
$$

The package does not use the usual chi-squared approximation to decide whether
this weighted statistic is large. It estimates a reference distribution by
parametric bootstrap.

## Step 4: calibrate the control limits

For each requested horizon $L$, one bootstrap replicate does the following:

1. Samples one candidate model using its weight $w_m$.
2. Draws stationary and invertible ARMA coefficients using the fitted
   coefficient covariance matrix.
3. Simulates $L$ in-control observations conditional on the reference history.
4. Joins the final $T-L$ reference observations to those $L$ simulated
   observations, keeping the window length equal to $T$.
5. Calculates $Q_{MM}(L)$ for that window.

The upper control limit is the selected quantile of the bootstrap statistics.
For example, `confidence_level = 0.95` uses the 95th percentile.

```{r calibrate}
calibration <- calibrate_qmm(
  fit,
  horizons = seq_along(simulated$monitoring),
  bootstrap_replicates = 99,
  confidence_level = 0.95,
  portmanteau_lag = 10,
  seed = 99
)

calibration_summary <- summary(calibration)
head(calibration_summary)
tail(calibration_summary, 3)
```

The diagnostic columns report the coefficient draws needed to obtain the 99
admissible bootstrap replicates at each horizon. Rejected draws are expected
when a sampled coefficient vector is nonstationary or noninvertible; unusually
many attempts deserve investigation.

Ninety-nine replicates keep this rendered vignette reasonably fast. They do
not make the estimated 95th percentiles precise enough for every application.
For an analysis used to make decisions, increase the number of replicates and
check whether the upper limits are sufficiently stable for the intended use.

## Step 5: monitor the new observations

At horizon $L$, the observed window contains the final $T-L$ reference
observations and the first $L$ new observations. A signal occurs when the
observed statistic is greater than its horizon-specific control limit.

```{r monitor}
monitoring <- monitor_qmm(calibration, simulated$monitoring)
monitoring

alerts <- signals(monitoring)
head(alerts, 8)
tail(alerts, 5)
```

The first signal in this fixed simulation occurs at Phase II horizon 4. Later
signals form a sustained cluster as observations from the changed process
replace more of the reference window. A single signal should trigger process
investigation; it does not prove that the controller change is the cause.

```{r plot, fig.width=7, fig.height=4.5}
plot(monitoring)
```

The dashed line is the control limit. Red points, if present, are signals.

For a `qcc`-style overview, align the complete series and the monitoring chart
on one time axis:

```{r overview-plot, fig.width=7, fig.height=6}
plot(monitoring, type = "overview")
```

The upper panel contains observations from Phase I and Phase II. The lower
panel contains the QMM statistic and its pointwise upper limits in Phase II.
Its shaded Phase I region has no retrospective QMM points because those
observations are used to fit and calibrate the method, not monitored against a
limit estimated from themselves.

### Inspect the bootstrap distributions behind the control limits

The control limits plotted above are quantiles of the horizon-specific
bootstrap distributions calculated in Step 4. `type = "bootstrap"` replaces
the shaded Phase I region with a row of density panels for a handful of those
distributions, so the reference sample's monitoring window and the bootstrap
calibration behind it can be read from a single figure:

```{r bootstrap-plot, fig.width=9, fig.height=6}
plot(monitoring, type = "bootstrap")
```

Each density panel shows one horizon's bootstrap `qmm_statistic()` values with
a dotted line at that horizon's control limit. By default, up to four horizons
already reached by monitoring are chosen automatically; pass `density_horizons`
to inspect specific ones instead, for example the first and last monitored
horizons:

```{r bootstrap-plot-chosen, fig.width=9, fig.height=6}
plot(monitoring, type = "bootstrap", density_horizons = c(1, 40))
```

`type = "bootstrap"` requires `keep_bootstrap = TRUE` in `calibrate_qmm()`
(the default), since it needs the retained bootstrap draws rather than only
the resulting limits.

### Add observations one at a time

A real application may receive one observation at a time. Calibrate the
planned horizons once, monitor the first new value, and then update the result.

```{r sequential}
live <- NULL
online_alerts <- list()

for (observation in simulated$monitoring) {
  live <- if (is.null(live)) {
    monitor_qmm(calibration, observation)
  } else {
    update(live, new_data = observation)
  }

  latest <- tail(summary(live), 1)

  if (latest$signal) {
    online_alerts[[length(online_alerts) + 1L]] <- latest
  }
}

online_alerts <- do.call(rbind, online_alerts)
row.names(online_alerts) <- NULL

head(online_alerts)
isTRUE(all.equal(online_alerts, alerts, check.attributes = FALSE))
```

`update()` uses the next limit already stored in `calibration`; it does not
refit the reference models or rerun the bootstrap. The calibration must
therefore contain every horizon that will be monitored. The final comparison
is `TRUE`: processing the observations one at a time produces the same alerts
as processing the complete Phase II vector.

## The one-call version

When one vector already contains both periods, `qmm_chart()` performs the same
three steps. `reference_length` specifies where monitoring begins.

```{r one-call, eval=FALSE}
chart <- qmm_chart(
  simulated$series,
  reference_length = 100,
  candidates = candidates,
  bootstrap_replicates = 199,
  portmanteau_lag = 10,
  seed = 99
)

chart
signals(chart)
plot(chart, type = "overview")
```

This block is not run while building the vignette because it repeats the fit
and bootstrap already demonstrated above. The larger bootstrap count is still
an example setting, not a universal requirement.

## Applying the workflow to your series

In practice, `x` can be any numeric vector or univariate `ts` object. For
example, after importing a CSV file whose `sensor_value` column is ordered by
collection time:

```{r own-series, eval=FALSE}
process_data <- read.csv("process-sensor.csv")
x <- process_data$sensor_value

reference_length <- 100

chart <- qmm_chart(
  x = x,
  reference_length = reference_length,
  candidates = arma_candidates(1, 1),
  bootstrap_replicates = 999,
  confidence_level = 0.95,
  portmanteau_lag = 10,
  seed = 2026
)

summary(chart$calibration$fit)
signals(chart)
plot(chart, type = "overview")
```

The initial `reference_length` observations must be a defensible in-control
reference period. The candidate orders, portmanteau lag, bootstrap size, and
confidence level are analysis choices; the values above illustrate the API
rather than replacing subject-matter justification.

## Difference from the original simulation prototype

The original prototype passed the true simulated innovations into the
bootstrap. True innovations are known in a simulation but are not observed in
a real process. The public package workflow instead uses the residual history
estimated by the candidate model selected in each bootstrap replicate.

This change makes the method usable with an observed series. It can also change
the numerical control limits. Simulation output still includes the true
innovations so that methodological validation can compare the feasible and
oracle versions without requiring them from package users.

Two other implementation differences affect the bootstrap. The original
prototype accepted coefficient draws without explicitly rejecting
nonstationary or noninvertible values; the package rejects those draws. Also,
`forecast::Arima()` in the prototype and `stats::arima()` in the package use
different conventions for the reported innovation variance even when their
coefficients, likelihood, AICc, weights, covariance matrices, and residuals
agree. These changes should be evaluated separately when comparing
implementations.

## Interpretation and scope

- Control limits are pointwise. Repeated monitoring changes the probability of
  seeing at least one false alarm over a complete run.
- The QMM construction supports horizons from 1 through $T$. It does not
  claim an indefinitely calibrated average run length.
- Results are conditional on the reference sample and candidate set.
- Model-fit and bootstrap diagnostics should be inspected before interpreting
  signals.

## References

For a package-based version of the original Monte Carlo experiment, see
`vignette("qmm-simulation-study", package = "fromage")`. For a guide to every
implementation helper and returned object, see
`vignette("qmm-internals", package = "fromage")`.
