---
title: "Reproducing the Original QMM Simulation Study"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Reproducing the Original QMM Simulation Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette rewrites the QMM part of the experiment in the original
`mmq_chart.Rmd` file using the public `fromage` API. It is a simulation study,
not the workflow required for monitoring one observed series. For that common
use case, start with `vignette("qmm-chart", package = "fromage")`.

The original notebook also compared QMM with a residual rule fitted by
`forecast::auto.arima()`. That rule is a separate benchmark and is not part of
the QMM method, so it is not reproduced here.

## What the experiment estimates

For one Phase II horizon (L), a Monte Carlo replicate performs these steps:

1. Simulate 100 Phase I observations from an ARMA(1,1) process.
2. Continue the same process for Phase II, possibly with changed coefficients.
3. Fit and weight the QMM candidate models using Phase I only.
4. Calibrate the pointwise upper control limit for horizon (L).
5. Form a length-100 coupled window with the last (100-L) Phase I
   observations and the first (L) Phase II observations.
6. Record whether its QMM statistic exceeds the limit.

When the Phase II coefficients equal the Phase I coefficients, the signal
proportion estimates the **pointwise false-alarm probability** at that horizon.
When the coefficients change, it estimates the **detection probability**, or
the power against that specified alternative at that horizon. These labels
should not be interchanged.

The intervals below are Wilson binomial-proportion intervals [@wilson1927].
They describe Monte Carlo uncertainty in the estimated signal proportion; they
are not QMM control limits.

## Experiment settings from the notebook

```{r settings}
library(fromage)

horizons <- c(1, 5, 15, 25, 50, 99)

phase1_coefficients <- c(ar1 = 0.4, ma1 = 0.3)
phase2_scenarios <- list(
  unchanged = c(ar1 = 0.4, ma1 = 0.3),
  moderate_increase = c(ar1 = 0.6, ma1 = 0.5),
  large_increase = c(ar1 = 0.8, ma1 = 0.7),
  decrease = c(ar1 = 0.2, ma1 = 0.1),
  sign_reversal = c(ar1 = -0.4, ma1 = -0.3)
)

candidates <- arma_candidates(max_p = 2, max_q = 2)
```

`arma_candidates(2, 2)` produces the eight ARMA orders used by the notebook:
all combinations from 0 through 2 except ARMA(0,0).

## Small helpers used only by the example

The following interval function implements the Wilson score formula. It is an
example helper, not part of the package API.

```{r wilson-helper}
wilson_interval <- function(successes, trials, confidence_level = 0.95) {
  z <- stats::qnorm(1 - (1 - confidence_level) / 2)
  estimate <- successes / trials
  denominator <- 1 + z^2 / trials
  center <- (estimate + z^2 / (2 * trials)) / denominator
  half_width <- z / denominator * sqrt(
    estimate * (1 - estimate) / trials + z^2 / (4 * trials^2)
  )

  c(
    estimate = estimate,
    lower = max(0, center - half_width),
    upper = min(1, center + half_width)
  )
}
```

At one horizon, the package need only calibrate that horizon. We therefore
calculate the observed coupled-window statistic directly with
`qmm_statistic()` instead of asking `monitor_qmm()` to evaluate every earlier
horizon as well.

```{r scenario-helper}
estimate_qmm_signals <- function(phase1_coefficients,
                                 phase2_coefficients,
                                 horizons,
                                 reference_length,
                                 monte_carlo_replicates,
                                 bootstrap_replicates,
                                 candidates,
                                 portmanteau_lag = 10,
                                 confidence_level = 0.95,
                                 burn_in = 1000) {
  rows <- vector("list", length(horizons))

  for (horizon_index in seq_along(horizons)) {
    horizon <- horizons[horizon_index]
    detected <- logical(monte_carlo_replicates)

    for (replicate in seq_len(monte_carlo_replicates)) {
      simulated <- simulate_arma_change(
        reference_length = reference_length,
        monitoring_length = reference_length,
        reference_coefficients = phase1_coefficients,
        monitoring_coefficients = phase2_coefficients,
        order = c(1, 1),
        burn_in = burn_in
      )

      fit <- fit_qmm(
        simulated$reference,
        candidates = candidates,
        criterion = "AICc"
      )

      calibration <- calibrate_qmm(
        fit,
        horizons = horizon,
        bootstrap_replicates = bootstrap_replicates,
        confidence_level = confidence_level,
        portmanteau_lag = portmanteau_lag,
        keep_bootstrap = FALSE
      )

      window <- if (horizon < reference_length) {
        c(
          simulated$reference[(horizon + 1):reference_length],
          simulated$monitoring[seq_len(horizon)]
        )
      } else {
        simulated$monitoring[seq_len(horizon)]
      }

      observed <- qmm_statistic(
        fit,
        window,
        portmanteau_lag = portmanteau_lag
      )
      detected[replicate] <- observed > calibration$limits$control_limit
    }

    interval <- wilson_interval(
      successes = sum(detected),
      trials = monte_carlo_replicates
    )
    rows[[horizon_index]] <- data.frame(
      horizon = horizon,
      signals = sum(detected),
      replicates = monte_carlo_replicates,
      estimate = unname(interval["estimate"]),
      lower = unname(interval["lower"]),
      upper = unname(interval["upper"])
    )
  }

  do.call(rbind, rows)
}
```

The outer helper repeats that calculation for every Phase II scenario. The
unchanged scenario is labelled separately from the alternatives.

```{r study-helper}
run_qmm_simulation_study <- function(phase1_coefficients,
                                     phase2_scenarios,
                                     horizons,
                                     reference_length = 100,
                                     monte_carlo_replicates = 1000,
                                     bootstrap_replicates = 500,
                                     candidates = arma_candidates(2, 2),
                                     portmanteau_lag = 10,
                                     confidence_level = 0.95,
                                     burn_in = 1000,
                                     seed = 2026) {
  set.seed(seed)
  scenario_rows <- vector("list", length(phase2_scenarios))

  for (scenario_index in seq_along(phase2_scenarios)) {
    coefficients <- phase2_scenarios[[scenario_index]]
    scenario_result <- estimate_qmm_signals(
      phase1_coefficients = phase1_coefficients,
      phase2_coefficients = coefficients,
      horizons = horizons,
      reference_length = reference_length,
      monte_carlo_replicates = monte_carlo_replicates,
      bootstrap_replicates = bootstrap_replicates,
      candidates = candidates,
      portmanteau_lag = portmanteau_lag,
      confidence_level = confidence_level,
      burn_in = burn_in
    )

    scenario_result$scenario <- names(phase2_scenarios)[scenario_index]
    unchanged <- isTRUE(all.equal(
      unname(coefficients),
      unname(phase1_coefficients)
    ))
    scenario_result$quantity <- if (unchanged) {
      "Pointwise false-alarm probability"
    } else {
      "Detection probability"
    }
    scenario_rows[[scenario_index]] <- scenario_result
  }

  result <- do.call(rbind, scenario_rows)
  rownames(result) <- NULL
  result[, c(
    "scenario", "quantity", "horizon", "signals", "replicates",
    "estimate", "lower", "upper"
  )]
}
```

## Fast executable example

The following call is deliberately tiny. Its purpose is to verify the complete
workflow quickly, not to provide stable Monte Carlo estimates.

```{r fast-study}
fast_results <- run_qmm_simulation_study(
  phase1_coefficients = phase1_coefficients,
  phase2_scenarios = phase2_scenarios[c("unchanged", "moderate_increase")],
  horizons = c(1, 5),
  reference_length = 40,
  monte_carlo_replicates = 2,
  bootstrap_replicates = 10,
  candidates = arma_candidates(1, 1),
  portmanteau_lag = 5,
  burn_in = 100,
  seed = 2026
)

fast_results
```

With only two Monte Carlo replicates and ten bootstrap replicates, these
numbers must not be interpreted as estimates of operating performance.

## Full notebook design

This call uses the values from `mmq_chart.Rmd`. It performs 30 combinations of
scenario and horizon, 1,000 Monte Carlo replicates per combination, and 500
bootstrap replicates within each Monte Carlo replicate. It is intentionally not
run while building the package because it is computationally expensive.

```{r full-study, eval=FALSE}
full_results <- run_qmm_simulation_study(
  phase1_coefficients = phase1_coefficients,
  phase2_scenarios = phase2_scenarios,
  horizons = horizons,
  reference_length = 100,
  monte_carlo_replicates = 1000,
  bootstrap_replicates = 500,
  candidates = arma_candidates(2, 2),
  portmanteau_lag = 10,
  confidence_level = 0.95,
  burn_in = 1000,
  seed = 2026
)

saveRDS(full_results, "qmm-simulation-study.rds")
```

Saving the result immediately avoids repeating a long experiment just to make
a table or plot.

```{r plot-full-study, eval=FALSE}
ggplot2::ggplot(
  full_results,
  ggplot2::aes(x = horizon, y = estimate, color = scenario, group = scenario)
) +
  ggplot2::geom_hline(yintercept = 0.05, linetype = "dashed") +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = lower, ymax = upper),
    width = 1,
    alpha = 0.5
  ) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~quantity) +
  ggplot2::labs(
    x = "Phase II horizon",
    y = "Estimated signal probability",
    color = "Phase II scenario"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

The horizontal 0.05 line is most directly relevant to the unchanged panel,
where it is the nominal pointwise false-alarm probability.

## Why the numbers can differ from the original notebook

The experiment design is the same, but the feasible package bootstrap does not
use the true simulated Phase I innovations. Those innovations are available to
the simulation author but are unobserved in a real application. Instead, each
package bootstrap replicate uses the finite residual history estimated by its
sampled Phase I candidate model.

This deliberate change makes the public procedure applicable to an observed
series. It also means the resulting control limits and signal proportions need
not be numerically identical to those produced by the original oracle-style
prototype.

The package also rejects nonstationary or noninvertible bootstrap coefficient
draws, whereas the original notebook did not explicitly reject them. In
addition, the original `forecast::Arima()` fit and the package's
`stats::arima()` fit report different innovation-variance estimates despite
agreement in coefficients, likelihood, AICc, weights, covariance matrices,
and residuals. Comparisons between implementations should introduce these
changes one at a time before comparing operating characteristics.

## References
